---
title: "Portland Data Science Group historical severe weather data visualization workshop"
output: html_notebook
---

```{r}

library(tidyverse)
library(fiftystater)
library(maps)
library(mapproj)
library(viridis)
library(ggmap)
library(plotly)

```

```{r}

# Portland Data Science Group data: https://drive.google.com/drive/folders/1JK1Oo6a85ZvEs1f3MM3wcGzCSAskXd-4
# NOAA website: https://www.ncdc.noaa.gov/stormevents/

# Jonathan Mackrory - event_reduced = broader groupings of events; also did some other cleanup

weather_trim3 <- read.csv("detail_trim3.csv")

```


```{r}

# number of rain events per state
# fiftystates needs lowercase state names
# plotly needs state abbreviations


weather_trim3 <- weather_trim3 %>%
  mutate(state = tolower(STATE))

state_abbs <- data.frame(STATE = toupper(state.name), state.abb)

weather_trim3 <- left_join(weather_trim3, state_abbs) 

```

# some queries don't cover all states (e.g. Rain events in 2016), so those need to be added as 0 or they will show as holes in the map 
# lookup table of 50 state names
# join with rain_2016_states
# replace NA with zero


statelist <- data.frame(state = tolower(state.name))

rain_2016_allstates <- left_join(statelist, rain_2016_states, by="state")

rain_2016_allstates$event_count[is.na(rain_2016_allstates$event_count)] <- 0



```{r}

# code from fiftystater vignette: https://cran.r-project.org/web/packages/fiftystater/vignettes/fiftystater.html


all_2016_ggplot <- weather_trim3 %>%
  filter(YEAR == 2016) %>%
  group_by(STATE, state, YEAR) %>%
  summarize(event_count = n()) %>%
  ggplot(aes(map_id = state)) +
  geom_map(aes(fill = event_count), map = fifty_states) + 
  expand_limits(x = fifty_states$long, y = fifty_states$lat) +
  coord_map() +
  scale_x_continuous(breaks = NULL) + 
  scale_y_continuous(breaks = NULL) +
  labs(x = "", y = "") +
  theme(legend.position = "bottom", 
        panel.background = element_blank()) +
  fifty_states_inset_boxes() +
  ggtitle("Number of severe weather events in 2016 by state") +
  scale_fill_viridis()

all_2016_ggplot

ggsave("all_weather.png", all_2016_ggplot)

```


```{r}

# have to filter data first, then do plotly (doesn't seem to work with filter piped directly into plotly)

# prepare columns adding indirect/direct injuries, indirect/direct deaths, property/crop damage

weather_trim3 <- weather_trim3 %>% 
  mutate(INJURIES_ALL = rowSums(weather_trim3[ ,14:15], na.rm = TRUE),
         DEATHS_ALL = rowSums(weather_trim3[ ,16:17], na.rm = TRUE),
         DAMAGE_ALL = rowSums(weather_trim3[ ,18:19], na.rm = TRUE))

# filter year and add number of event and sums of injuries, deaths, damage

plotly_2016_data <- weather_trim3 %>%
  filter(YEAR == 2016) %>%
  group_by(state.abb) %>%
  summarize(event_count = n(), Injuries_total = sum(INJURIES_ALL), Deaths_total = sum(DEATHS_ALL), Damage_total = sum(DAMAGE_ALL))

# add hover column for mouse-over text
# use prettyNum to add commas to dollar amounts

plotly_2016_data$hover <- c(paste0(plotly_2016_data$state.abb,
                                   "\nInjuries = ", plotly_2016_data$Injuries_total, 
                                "\nDeaths = ", plotly_2016_data$Deaths_total, 
                                "\nDamage = $", prettyNum(plotly_2016_data$Damage_total, big.mark = ",", scientific = FALSE, trim = TRUE)))

```

```{r}

plotly_2016 <- plot_ly(plotly_2016_data, type = "choropleth", 
        locations = plotly_2016_data$state.abb,
        locationmode = "USA-states",
        z = plotly_2016_data$event_count, text = plotly_2016_data$hover) %>%
  layout(geo = list(scope = "usa"), title = "All weather events in 2016 by state")

htmlwidgets::saveWidget(plotly_2016, "plotly_2016_weather.html")

## Next step - make Shiny app with years, other info

```






















